---
title: "Optiver Baseline in Ibex + R"
output:
  html_document:
    toc: true
    toc_depth: 2
params:
  train_csv: "../../kaggle/data/train.csv"
  max_date_id: 40
  sample_rows: 250000
  n_folds: 5
  lightbm_iterations: 300
  lightbm_learning_rate: 0.03
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Goal

This R Markdown document translates the local Kaggle notebook
`baseline-lgb-xgb-and-catboost.ipynb` into a pure R workflow that:

- uses `ribex` to stage the Optiver CSV to Parquet once, then reuse Parquet on later runs
- keeps the notebook's pairwise and triple-price imbalance features
- fits the Ibex `lightbm` plugin on each fold
- scores the held-out folds in R from the returned coefficient table

The original Python notebook defaults to inference with pretrained models.
Here, the R translation is set up for local training instead.

Render from the repo root with:

```sh
./scripts/render-optiver-baseline-rmd.sh
```

Or choose another `rmarkdown` output format explicitly:

```sh
./scripts/render-optiver-baseline-rmd.sh pdf_document
```

## Packages and Paths

```{r packages}
library(ribex)
library(ggplot2)

ribex::register_knitr_engines()

resolve_repo_root <- function() {
  candidates <- c(getwd(), normalizePath("..", winslash = "/", mustWork = FALSE))
  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "r", "ribex", "DESCRIPTION"))) {
      return(normalizePath(candidate, winslash = "/", mustWork = TRUE))
    }
  }
  stop("Could not locate the Ibex repo root from the current working directory.")
}

repo_root <- resolve_repo_root()
notebook_dir <- file.path(repo_root, "notebooks")
train_csv <- normalizePath(file.path(notebook_dir, params$train_csv), winslash = "/", mustWork = TRUE)
plugin_paths <- ribex::default_plugin_paths()
cache_dir <- file.path(notebook_dir, "_cache")
train_parquet <- file.path(cache_dir, "optiver-train.parquet")

if (!length(plugin_paths)) {
  stop("Could not find any Ibex plugin directories. Set IBEX_BUILD_DIR or IBEX_LIBRARY_PATH.")
}

list(
  repo_root = repo_root,
  train_csv = train_csv,
  train_parquet = train_parquet,
  plugin_paths = plugin_paths,
  lightbm_iterations = params$lightbm_iterations,
  lightbm_learning_rate = params$lightbm_learning_rate
)
```

## Stage CSV to Parquet and Load Once in an Ibex Session

```{r stage}
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

escape_ibex_string <- function(x) {
  x <- gsub("\\\\", "\\\\\\\\", x)
  gsub('"', '\\"', x, fixed = TRUE)
}

ensure_optiver_parquet <- function(train_csv, train_parquet, plugin_paths, repo_root) {
  if (file.exists(train_parquet)) {
    return(invisible(train_parquet))
  }

  ibex_bin <- file.path(repo_root, "build-release", "tools", "ibex")
  if (!file.exists(ibex_bin)) {
    stop("Could not find the Ibex REPL binary at ", ibex_bin)
  }

  stage_script <- tempfile(pattern = "optiver_stage_", fileext = ".ibex")
  on.exit(unlink(stage_script), add = TRUE)

  script_lines <- c(
    'extern fn read_csv(path: String, nulls: String) -> DataFrame from "csv.hpp";',
    'extern fn write_parquet(df: DataFrame, path: String) -> Int from "parquet.hpp";',
    sprintf('let train = read_csv("%s", "<empty>");', escape_ibex_string(train_csv)),
    sprintf('write_parquet(train, "%s");', escape_ibex_string(train_parquet)),
    'train[select { rows = count() }];'
  )
  writeLines(script_lines, stage_script)

  ibex_env <- c(sprintf("IBEX_LIBRARY_PATH=%s", paste(plugin_paths, collapse = .Platform$path.sep)))
  out <- system2(
    ibex_bin,
    args = c("--plugin-path", plugin_paths[[1L]]),
    input = c(sprintf(":load %s", stage_script), ":q"),
    stdout = TRUE,
    stderr = TRUE,
    env = ibex_env
  )

  status <- attr(out, "status")
  if (!is.null(status) && status != 0) {
    stop("Ibex Parquet staging failed:\n", paste(out, collapse = "\n"))
  }
  if (!file.exists(train_parquet)) {
    stop("Ibex Parquet staging did not create ", train_parquet,
         "\nOutput:\n", paste(out, collapse = "\n"))
  }

  invisible(train_parquet)
}

ensure_optiver_parquet(train_csv, train_parquet, plugin_paths, repo_root)

sess <- create_session(plugin_paths = plugin_paths)
```

```{ibex load-train, session="optiver", scalars=list(train_parquet=train_parquet), quiet=TRUE}
extern fn read_parquet(path: String) -> DataFrame from "parquet.hpp";
let train = read_parquet(train_parquet);
```

```{r define-train-work}
if (!is.null(params$max_date_id)) {
  session_eval(
    ribex::knitr_session("optiver", plugin_paths),
    "
    let train_work = train[filter date_id <= max_date_id];
    ",
    scalars = list(max_date_id = as.integer(params$max_date_id))
  )
} else {
  session_eval(ribex::knitr_session("optiver", plugin_paths), "let train_work = train;")
}
```

```{ibex train-rows, session="optiver", assign="train_rows"}
train_work[select { rows = count() }];
```

## Define the Feature Families

```{r feature-query}
price_cols <- c("reference_price", "far_price", "near_price", "ask_price", "bid_price", "wap")

base_features <- c(
  "seconds_in_bucket",
  "imbalance_buy_sell_flag",
  "imbalance_size",
  "matched_size",
  "bid_size",
  "ask_size",
  "reference_price",
  "far_price",
  "near_price",
  "ask_price",
  "bid_price",
  "wap",
  "imb_s1",
  "imb_s2"
)

pair_feature_names <- apply(
  combn(price_cols, 2),
  2,
  \(cols) sprintf("%s_%s_imb", cols[[2]], cols[[1]])
)

triple_feature_names <- apply(
  combn(price_cols, 3),
  2,
  \(cols) sprintf("%s_%s_%s_imb2", cols[[3]], cols[[2]], cols[[1]])
)

feature_cols <- c(
  base_features,
  pair_feature_names,
  triple_feature_names
)

length(feature_cols)
```

## Materialize the Feature Table

```{ibex features, session="optiver", assign="train_features", quiet=TRUE}
let price_cols = ["reference_price", "far_price", "near_price", "ask_price", "bid_price", "wap"];

train_work[update {
    imb_s1 = (bid_size - ask_size) / (bid_size + ask_size),
    imb_s2 = (imbalance_size - matched_size) / (matched_size + imbalance_size),
    map (i, a) in price_cols, (j, b) in price_cols
        where i > j
        => `${a}_${b}_imb` = (get(a) - get(b)) / (get(a) + get(b)),
    map (i, a) in price_cols, (j, b) in price_cols, (k, c) in price_cols
        where i > j && j > k
        => `${a}_${b}_${c}_imb2` =
            (
                pmax(get(a), get(b), get(c)) -
                (get(a) + get(b) + get(c) - pmin(get(a), get(b), get(c)) - pmax(get(a), get(b), get(c)))
            ) /
            (
                (get(a) + get(b) + get(c) - pmin(get(a), get(b), get(c)) - pmax(get(a), get(b), get(c))) -
                pmin(get(a), get(b), get(c))
            )
}];
```

```{ibex features-rows, session="optiver"}
train_features[select { rows = count() }];
```

```{ibex features-preview, session="optiver"}
train_features[select {
    stock_id,
    date_id,
    seconds_in_bucket,
    target,
    reference_price,
    far_price,
    near_price,
    ask_price,
    bid_price,
    wap,
    imb_s1,
    imb_s2
}];
```

## Prepare the Training Matrix

```{r training-data}
model_df <- train_features[is.finite(train_features$target), c("row_id", "target", feature_cols)]

if (!is.null(params$sample_rows) && nrow(model_df) > params$sample_rows) {
  set.seed(42)
  take <- sample.int(nrow(model_df), size = params$sample_rows)
  model_df <- model_df[take, , drop = FALSE]
}

feature_matrix_df <- model_df[, feature_cols, drop = FALSE]
feature_matrix_df[] <- lapply(feature_matrix_df, function(col) {
  col <- as.numeric(col)
  col[!is.finite(col)] <- NA_real_
  fill <- stats::median(col, na.rm = TRUE)
  if (!is.finite(fill)) {
    fill <- 0.0
  }
  col[is.na(col)] <- fill
  col
})

model_input_df <- cbind(
  model_df[, c("row_id", "target"), drop = FALSE],
  feature_matrix_df
)

fold_id <- (seq_len(nrow(model_df)) - 1L) %% params$n_folds
X <- as.matrix(feature_matrix_df)
storage.mode(X) <- "double"
y <- model_df$target

list(
  rows = nrow(model_input_df),
  cols = ncol(X),
  folds = params$n_folds,
  feature_median_imputed = mean(!is.finite(as.matrix(model_df[, feature_cols, drop = FALSE])) | is.na(as.matrix(model_df[, feature_cols, drop = FALSE])))
)
```

## Train the Ibex LightBM Baseline

```{r train-models}
mae <- function(actual, predicted) {
  mean(abs(actual - predicted))
}

build_lightbm_query <- function(feature_cols) {
  feature_formula <- paste(feature_cols, collapse = " + ")
  paste(
    'extern fn lightbm_version() -> DataFrame from "lightbm.hpp";',
    sprintf(
      "train_fold[model { target ~ %s, method = lightbm, iterations = iterations, learning_rate = learning_rate }];",
      feature_formula
    )
  )
}

predict_from_coefficients <- function(coef_table, new_data, feature_cols) {
  beta <- stats::setNames(coef_table$estimate, coef_table$term)
  intercept <- if ("(intercept)" %in% names(beta)) beta[["(intercept)"]] else 0.0
  feature_beta <- beta[feature_cols]
  if (anyNA(feature_beta)) {
    missing_terms <- feature_cols[is.na(feature_beta)]
    stop("Missing coefficients for features: ", paste(missing_terms, collapse = ", "))
  }
  intercept + as.numeric(as.matrix(new_data[, feature_cols, drop = FALSE]) %*% feature_beta)
}

fit_ibex_lightbm <- function(model_input_df, fold_id, n_folds, feature_cols, plugin_paths,
                             iterations, learning_rate) {
  oof <- rep(NA_real_, nrow(model_input_df))
  coefficients <- vector("list", n_folds)
  sess <- create_session(plugin_paths = plugin_paths)
  model_query <- build_lightbm_query(feature_cols)

  for (i in seq_len(n_folds) - 1L) {
    train_idx <- fold_id != i
    valid_idx <- fold_id == i
    train_fold <- model_input_df[train_idx, c("target", feature_cols), drop = FALSE]
    valid_fold <- model_input_df[valid_idx, feature_cols, drop = FALSE]

    coefficients[[i + 1L]] <- session_eval(
      sess,
      model_query,
      tables = list(train_fold = train_fold),
      scalars = list(
        iterations = as.integer(iterations),
        learning_rate = as.numeric(learning_rate)
      ),
      format = "data.frame"
    )

    oof[valid_idx] <- predict_from_coefficients(
      coefficients[[i + 1L]],
      valid_fold,
      feature_cols
    )
  }

  list(
    name = "ibex_lightbm",
    coefficients = coefficients,
    oof = oof,
    mae = mae(model_input_df$target, oof)
  )
}

model_result <- fit_ibex_lightbm(
  model_input_df = model_input_df,
  fold_id = fold_id,
  n_folds = params$n_folds,
  feature_cols = feature_cols,
  plugin_paths = plugin_paths,
  iterations = params$lightbm_iterations,
  learning_rate = params$lightbm_learning_rate
)

metrics <- data.frame(model = model_result$name, mae = model_result$mae)
metrics
```

## Diagnostics

```{r diagnostics}
print(metrics)

coef_long <- do.call(
  rbind,
  lapply(seq_along(model_result$coefficients), function(i) {
    transform(model_result$coefficients[[i]], fold = i)
  })
)

coef_summary <- aggregate(
  abs(estimate) ~ term,
  data = subset(coef_long, term != "(intercept)"),
  FUN = mean
)
coef_summary <- coef_summary[order(coef_summary$`abs(estimate)`, decreasing = TRUE), , drop = FALSE]
top_coefficients <- head(coef_summary, 20L)

coef_plot <- ggplot(top_coefficients, aes(x = reorder(term, `abs(estimate)`), y = `abs(estimate)`)) +
  geom_col(fill = "#3b7ea1") +
  coord_flip() +
  labs(
    title = "Top average absolute LightBM coefficients across folds",
    x = NULL,
    y = "Mean |estimate|"
  ) +
  theme_minimal(base_size = 12)

print(coef_plot)

sample_n <- min(5000L, length(y))
set.seed(42)
idx <- sample.int(length(y), sample_n)
scatter_df <- data.frame(actual = y[idx], predicted = model_result$oof[idx])

scatter_plot <- ggplot(scatter_df, aes(actual, predicted)) +
  geom_point(alpha = 0.25, size = 0.7, color = "#1f4e5f") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#b22222") +
  labs(
    title = "Ibex LightBM predictions vs actual target",
    x = "Actual target",
    y = "Predicted target"
  ) +
  theme_minimal(base_size = 12)

print(scatter_plot)
```

## Notes

- The original notebook used pretrained models for Kaggle inference. This R version is set up for local training instead.
- The expensive part remains `read_csv(...)`; using an Ibex session ensures it happens once per knit.
- The feature engineering stays close to the notebook logic and now runs fully through the Ibex feature query before the model handoff to R.
- If you want a full-size run, increase `params$max_date_id` or set it to `NULL`, and increase `params$sample_rows`.
