---
title: "Optiver Baseline in Ibex + R"
output:
  html_document:
    toc: true
    toc_depth: 2
params:
  train_csv: "../../kaggle/data/train.csv"
  max_date_id: 40
  sample_rows: 250000
  n_folds: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Goal

This R Markdown document translates the local Kaggle notebook
`baseline-lgb-xgb-and-catboost.ipynb` into a pure R workflow that:

- uses `ribex` to load the Optiver CSV once and build the feature table
- keeps the notebook's pairwise and triple-price imbalance features
- trains whichever of `lightgbm`, `xgboost`, and `catboost` are available
- evaluates them with a simple fold split and an equal-weight ensemble

The original Python notebook defaults to inference with pretrained models.
Here, the R translation is set up for local training instead.

Render from the repo root with:

```sh
./scripts/render-optiver-baseline-rmd.sh
```

Or choose another `rmarkdown` output format explicitly:

```sh
./scripts/render-optiver-baseline-rmd.sh pdf_document
```

## Packages and Paths

```{r packages}
library(ribex)
library(ggplot2)

resolve_repo_root <- function() {
  candidates <- c(getwd(), normalizePath("..", winslash = "/", mustWork = FALSE))
  for (candidate in unique(candidates)) {
    if (file.exists(file.path(candidate, "r", "ribex", "DESCRIPTION"))) {
      return(normalizePath(candidate, winslash = "/", mustWork = TRUE))
    }
  }
  stop("Could not locate the Ibex repo root from the current working directory.")
}

repo_root <- resolve_repo_root()
notebook_dir <- file.path(repo_root, "notebooks")
train_csv <- normalizePath(file.path(notebook_dir, params$train_csv), winslash = "/", mustWork = TRUE)
plugin_paths <- ribex::default_plugin_paths()

if (!length(plugin_paths)) {
  stop("Could not find any Ibex plugin directories. Set IBEX_BUILD_DIR or IBEX_LIBRARY_PATH.")
}

optional_models <- c("lightgbm", "xgboost", "catboost")
available_models <- optional_models[vapply(optional_models, requireNamespace, logical(1), quietly = TRUE)]

list(
  repo_root = repo_root,
  train_csv = train_csv,
  plugin_paths = plugin_paths,
  available_models = available_models
)
```

## Load the CSV Once in an Ibex Session

```{r load-train}
sess <- create_session(plugin_paths = plugin_paths)

session_eval(
  sess,
  '
  extern fn read_csv(path: String, nulls: String) -> DataFrame from "csv.hpp";
  let train = read_csv(train_csv, "<empty>");
  ',
  scalars = list(train_csv = train_csv)
)

if (!is.null(params$max_date_id)) {
  session_eval(
    sess,
    "
    let train_work = train[filter date_id <= max_date_id];
    ",
    scalars = list(max_date_id = as.integer(params$max_date_id))
  )
} else {
  session_eval(sess, "let train_work = train;")
}

train_rows <- session_eval(sess, "train_work[select { rows = count() }];")
train_rows
```

## Generate the Ibex Feature Query

```{r feature-query}
price_cols <- c("reference_price", "far_price", "near_price", "ask_price", "bid_price", "wap")

base_features <- c(
  "seconds_in_bucket",
  "imbalance_buy_sell_flag",
  "imbalance_size",
  "matched_size",
  "bid_size",
  "ask_size",
  "reference_price",
  "far_price",
  "near_price",
  "ask_price",
  "bid_price",
  "wap",
  "imb_s1",
  "imb_s2"
)

safe_div <- function(num, den) sprintf("((%s) / (%s))", num, den)

pair_feature_specs <- list()
for (i in seq_along(price_cols)) {
  for (j in seq_along(price_cols)) {
    if (i > j) {
      a <- price_cols[[i]]
      b <- price_cols[[j]]
      name <- sprintf("%s_%s_imb", a, b)
      expr <- safe_div(sprintf("(%s - %s)", a, b), sprintf("(%s + %s)", a, b))
      pair_feature_specs[[length(pair_feature_specs) + 1L]] <- list(name = name, expr = expr)
    }
  }
}

triple_feature_names <- c()
for (i in seq_along(price_cols)) {
  for (j in seq_along(price_cols)) {
    for (k in seq_along(price_cols)) {
      if (i > j && j > k) {
        triple_feature_names <- c(
          triple_feature_names,
          sprintf("%s_%s_%s_imb2", price_cols[[i]], price_cols[[j]], price_cols[[k]])
        )
      }
    }
  }
}

derived_assignments <- c(
  sprintf("imb_s1 = %s", safe_div("(bid_size - ask_size)", "(bid_size + ask_size)")),
  sprintf("imb_s2 = %s", safe_div("(imbalance_size - matched_size)", "(matched_size + imbalance_size)")),
  vapply(pair_feature_specs, function(spec) sprintf("%s = %s", spec$name, spec$expr), character(1))
)

feature_cols <- c(
  base_features,
  vapply(pair_feature_specs, `[[`, character(1), "name"),
  triple_feature_names
)

build_feature_query <- function(table_name = "train_work") {
  paste0(
    table_name,
    "[update {\n",
    paste0("  ", derived_assignments, collapse = ",\n"),
    "\n}];"
  )
}

feature_query <- build_feature_query()
cat(substr(feature_query, 1L, 1500L))
if (nchar(feature_query) > 1500L) {
  cat("\n...\n")
}

length(feature_cols)
```

## Materialize the Feature Table

```{r features, cache=TRUE}
train_features <- session_eval(sess, feature_query, format = "data.frame")

dim(train_features)
head(train_features[, c("stock_id", "date_id", "seconds_in_bucket", "target", feature_cols[1:8])])
```

## Add the Triple-Price Imbalance Features in R

The original notebook computes the 3-price imbalance terms row-wise with
`max`, `min`, and `mid`. The current interpreted `ribex` session runtime still
lacks the scalar `abs` / `sqrt` support needed to express the same row-wise
formula entirely inside the notebook session, so this final feature family is
added in R after the Ibex result comes back.

```{r triple-features}
add_triple_features <- function(df, prices) {
  out <- df
  for (i in seq_along(prices)) {
    for (j in seq_along(prices)) {
      for (k in seq_along(prices)) {
        if (i > j && j > k) {
          a <- prices[[i]]
          b <- prices[[j]]
          c <- prices[[k]]
          max_ <- pmax(out[[a]], out[[b]], out[[c]], na.rm = FALSE)
          min_ <- pmin(out[[a]], out[[b]], out[[c]], na.rm = FALSE)
          mid_ <- out[[a]] + out[[b]] + out[[c]] - max_ - min_
          name <- sprintf("%s_%s_%s_imb2", a, b, c)
          out[[name]] <- (max_ - mid_) / (mid_ - min_)
        }
      }
    }
  }
  out
}

train_features <- add_triple_features(train_features, price_cols)

head(train_features[, c("wap_bid_price_ask_price_imb2", "wap_bid_price_near_price_imb2")])
```

## Prepare the Training Matrix

```{r training-data}
model_df <- train_features[is.finite(train_features$target), c("row_id", "target", feature_cols)]

if (!is.null(params$sample_rows) && nrow(model_df) > params$sample_rows) {
  set.seed(42)
  take <- sample.int(nrow(model_df), size = params$sample_rows)
  model_df <- model_df[take, , drop = FALSE]
}

fold_id <- (seq_len(nrow(model_df)) - 1L) %% params$n_folds
X <- as.matrix(model_df[, feature_cols, drop = FALSE])
y <- model_df$target

list(
  rows = nrow(model_df),
  cols = ncol(X),
  folds = params$n_folds,
  missing_share = mean(!is.finite(X) | is.na(X))
)
```

## Train Optional Baseline Models

```{r train-models}
mae <- function(actual, predicted) {
  mean(abs(actual - predicted))
}

fit_lightgbm <- function(X, y, fold_id, n_folds) {
  oof <- rep(NA_real_, length(y))
  models <- vector("list", n_folds)
  for (i in seq_len(n_folds) - 1L) {
    train_idx <- fold_id != i
    valid_idx <- fold_id == i
    dtrain <- lightgbm::lgb.Dataset(X[train_idx, , drop = FALSE], label = y[train_idx])
    dvalid <- lightgbm::lgb.Dataset(X[valid_idx, , drop = FALSE], label = y[valid_idx])
    models[[i + 1L]] <- lightgbm::lgb.train(
      params = list(objective = "regression_l1", metric = "l1"),
      data = dtrain,
      valids = list(valid = dvalid),
      nrounds = 500,
      early_stopping_rounds = 100,
      verbose = 10
    )
    oof[valid_idx] <- predict(models[[i + 1L]], X[valid_idx, , drop = FALSE])
  }
  list(name = "lightgbm", models = models, oof = oof, mae = mae(y, oof))
}

fit_xgboost <- function(X, y, fold_id, n_folds) {
  oof <- rep(NA_real_, length(y))
  models <- vector("list", n_folds)
  for (i in seq_len(n_folds) - 1L) {
    train_idx <- fold_id != i
    valid_idx <- fold_id == i
    dtrain <- xgboost::xgb.DMatrix(X[train_idx, , drop = FALSE], label = y[train_idx], missing = NA)
    dvalid <- xgboost::xgb.DMatrix(X[valid_idx, , drop = FALSE], label = y[valid_idx], missing = NA)
    models[[i + 1L]] <- xgboost::xgb.train(
      params = list(
        objective = "reg:absoluteerror",
        eval_metric = "mae",
        tree_method = "hist"
      ),
      data = dtrain,
      watchlist = list(valid = dvalid),
      nrounds = 500,
      early_stopping_rounds = 100,
      verbose = 10
    )
    oof[valid_idx] <- predict(models[[i + 1L]], dvalid)
  }
  list(name = "xgboost", models = models, oof = oof, mae = mae(y, oof))
}

fit_catboost <- function(X, y, fold_id, n_folds) {
  oof <- rep(NA_real_, length(y))
  models <- vector("list", n_folds)
  for (i in seq_len(n_folds) - 1L) {
    train_idx <- fold_id != i
    valid_idx <- fold_id == i
    train_pool <- catboost::catboost.load_pool(X[train_idx, , drop = FALSE], label = y[train_idx])
    valid_pool <- catboost::catboost.load_pool(X[valid_idx, , drop = FALSE], label = y[valid_idx])
    models[[i + 1L]] <- catboost::catboost.train(
      learn_pool = train_pool,
      test_pool = valid_pool,
      params = list(
        loss_function = "MAE",
        iterations = 3000,
        verbose = 100
      )
    )
    oof[valid_idx] <- catboost::catboost.predict(models[[i + 1L]], valid_pool)
  }
  list(name = "catboost", models = models, oof = oof, mae = mae(y, oof))
}

model_results <- list()

if ("lightgbm" %in% available_models) {
  model_results[["lightgbm"]] <- fit_lightgbm(X, y, fold_id, params$n_folds)
}
if ("xgboost" %in% available_models) {
  model_results[["xgboost"]] <- fit_xgboost(X, y, fold_id, params$n_folds)
}
if ("catboost" %in% available_models) {
  model_results[["catboost"]] <- fit_catboost(X, y, fold_id, params$n_folds)
}

if (!length(model_results)) {
  message("No optional model packages are installed. Install one or more of: ",
          paste(optional_models, collapse = ", "))
}

metrics <- if (length(model_results)) {
  data.frame(
    model = vapply(model_results, `[[`, character(1), "name"),
    mae = vapply(model_results, `[[`, numeric(1), "mae")
  )
} else {
  data.frame(model = character(), mae = numeric())
}

metrics
```

## Ensemble and Diagnostics

```{r diagnostics}
if (length(model_results)) {
  pred_matrix <- do.call(cbind, lapply(model_results, `[[`, "oof"))
  ensemble_pred <- rowMeans(pred_matrix)
  ensemble_mae <- mae(y, ensemble_pred)

  metrics <- rbind(
    metrics,
    data.frame(model = "ensemble_mean", mae = ensemble_mae)
  )

  print(metrics[order(metrics$mae), , drop = FALSE])

  metrics_plot <- ggplot(metrics, aes(x = reorder(model, mae), y = mae)) +
    geom_col(fill = "#3b7ea1") +
    coord_flip() +
    labs(
      title = "Fold MAE by model",
      x = NULL,
      y = "MAE"
    ) +
    theme_minimal(base_size = 12)

  print(metrics_plot)

  sample_n <- min(5000L, length(y))
  set.seed(42)
  idx <- sample.int(length(y), sample_n)
  scatter_df <- data.frame(actual = y[idx], predicted = ensemble_pred[idx])

  scatter_plot <- ggplot(scatter_df, aes(actual, predicted)) +
    geom_point(alpha = 0.25, size = 0.7, color = "#1f4e5f") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#b22222") +
    labs(
      title = "Ensemble predictions vs actual target",
      x = "Actual target",
      y = "Predicted target"
    ) +
    theme_minimal(base_size = 12)

  print(scatter_plot)
}
```

## Notes

- The original notebook used pretrained models for Kaggle inference. This R version is set up for local training instead.
- The expensive part remains `read_csv(...)`; using an Ibex session ensures it happens once per knit.
- The feature engineering stays close to the notebook logic. Import, the core features, and the pairwise price imbalance terms run in Ibex; the 3-price imbalance terms are added in R because the current interpreted runtime still lacks row-wise scalar `abs` / `sqrt`.
- If you want a full-size run, increase `params$max_date_id` or set it to `NULL`, and increase `params$sample_rows`.
